Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-23 latex/main.tex
Decent pass on first para

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with 
% Not sure about this assertion! Yan?
samples separated by millions of years of evolution.
[This is rough outline text]
Complex models are required to fill in the gaps in our knowledge,
incorporating prior information about sequence evoltion etc etc.
In this context, recombination can be treated as a nuisance factor 
because we cannot hope to observe it over these timescales [or something
similar].
Phylogenetics has continued along this vein, 
although with various challenges on how to accomodate 
more closely related samples and to handle potential recombination met.

Recombination breaks things\cite{Posada2002effect}

ARG/phylonetwork -like
methods~\cite{Didelot2010inference,Vaughan2017inferring,
Muller2020bayesian,Muller2022bayesian}

Distinction between recombination detection --- identifying that a 
given sequences is a likely recombinant --- and incorporation
of recombination.

Recombination has been incorporated in a number of models

>>>>

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with small numbers of widely diverged
samples. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of samples from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches 
used to perform species-level tree 
inference~\cite{Suchard2018bayesian,Bouckaert2019beast},
comprehensive phylogenies of these datasets are typically
built with simpler distance-matrix methods [citation?]
or approximations~\cite{Price2009fasttree}, primarily
due to computational scaling limitations, but also because of
complexity of modelling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic 
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Muller2022bayesian}, 
scalability 
is a major issue and practical application has been limited.
